Sampling Strategies in Decisions from Experience

### 2022-02-06 manuscript.Rmd
Change Figure 1

Since there is an uncountable number of decisions people confront in the natural world, behavioral decision research routinely abstracts from decisions between particular choice alternatives, e.g., the choice between job offers, political parties to vote for, travel destinations, investment plans, and whatnot.
Rather, with the choice between at least two *prospects* (see below), it studies a case which is counterfactual in that it omits the many particularities of each decision, but retains in the form of prospects the fundamental *properties* that almost all choice alternatives are assumed to possess [@lopesThoughtsPsychologicalConcept1983].^[See @knightRiskUncertaintyProfit1921, for a discussion of the reasonableness of such abstractions.] 
These fundamental properties are considered to be the possible outcomes of a choice alternative---hereafter denoted by the Greek letter $\omega$---and the probabilities with which these outcomes occur following the choice of the alternative---hereafter $p(\omega)$.
That is, any choice alternative that can be fully described by its outcomes and their probabilities is a prospect. 
For illustration, take the choice alternatives in the following example with monetary outcomes:

@@

Now, let the exact axiomatic definition of a prospect be that of a *probability space* denoted by the tuple $(\Omega, \mathcal{F}, P)$. 
Following the @kolmogorovFoundationsTheoryProbability1950 axioms, $\Omega$ denotes the sample space, containing a finite set of possible outcomes:
$$
\omega_i \in \{\omega_1, ..., \omega_n\} = \Omega
\; .
\tag{1}
$$ 
$\mathcal{F}$ is the set of all possible subsets of $\Omega$, i.e., the power set $\mathcal{P}(\Omega)$, with each element $A \in P(\Omega)$ being a subset of $\Omega$.^[Note that this paper considers only the case of finite sample spaces. The exact axiomatic definition of $\mathcal{F}$, which can be found in Kolmogorov [-@kolmogorovFoundationsTheoryProbability1950, pp. 2-3 and 14-15], also covers infinite sample spaces and is that of a $\sigma$-algebra. For the finite case, however, $\mathcal{P}(\Omega)$ can be considered as such [@georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.1].] 
$P$ is a probability measure
$$
P: \mathcal{F} \mapsto [0,1]
\; ,
\tag{2}
$$
which assigns each subset $A \in \mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.^[The probabilities $p(\omega)$ are determined by the probability mass function $p : \Omega \mapsto [0,1]$, which is used to construct the probability measure $P(A) = \sum_{\omega \in A} p(\omega)$ for all $A \in \mathcal{F}$.]
Given a choice between 1, ..., k prospects, the inputs from the environment or memory should therefore provide information about the set
$$
I = \{(\Omega_1, \mathcal{F}_1, P_1), ... ,(\Omega_k, \mathcal{F}_k, P_k)\}
\; .
\tag{3}
$$

However, rarely in peoples' daily life, inputs take the form of explicit descriptions of all outcomes and their probabilities (as in Example 1).
In such cases, people would make *decisions from description* (DfD), since the information from which the mind learns about the choice alternatives---the probability spaces in (3)---are complete descriptions thereof.
Yet, rather often, people make *decisions from experience* [DFE, @hertwigDecisionsExperienceEffect2004], where the mind can learn about the choice alternatives only from a record of past experiences with them, which necessarily provides only incomplete and probable information.
That is, people may have experienced the possible outcomes of a choice alternative when they themselves or others have made the same decision repeatedly in the past, and the record of experienced outcomes for this particular choice alternative is denoted a *sample*. 
The mind may then infer from the relative frequency with which an outcome occurs in such a sample of past decisions its probability of occurrence for a future decision. 
Of course, all other things being equal, the accuracy of this inference depends on the size of the sample.
<!--For illustration, take the sampled outcomes in the following example, where alternative A2 from Example 1 was chosen ten times without prior knowledge about the possible outcomes of this alternative and their probabilities:

@@

\noindent
One may now make an inference from the sample about the possible outcomes of choice alternative A2 and their probabilities. 
However, this conclusion is not certain, but only probable, since the relative frequencies of the results will almost surely change as the sample size increases.-->

@@

The choices people actually make between prospects---be that on the basis of ---are often described in terms of deviations from expected value (EV) maximization, where the EV is given by 
$$
EV = \sum_i^n \omega_i \times P(X = \omega_i)
\; .
\tag{6}
$$ 
That is, following the principle of EV maximization, one would first calculate the prospects' EV according to (6) and thereafter chose the prospect with the largest EV.
To *describe* how peoples' choices deviate from EV maximization [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations], the models of cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992; see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data with a combination of two parameterized functions that may replace the objective outcomes and probabilities in (3) by nonlinear transformations thereof.^[Note that CPT or other originally deterministic decision models may be amended by a stochastic choice function [see, e.g., @rieskampProbabilisticNaturePreferential2008].] 
That is, the models of CPT fit choice data by maximizing the value 
$$
V = \sum_i^n v(\omega_i) \times \pi_i
\; ,
\tag{5}
$$ 
with subjective values $v(\omega)$ derived from a *value function* that may transform the objective outcomes, and decision weights $\pi$ derived from the objective probabilities' cumulative distribution function that may be transformed by a *probability weighting function* (see below). 
Now, since the probability weighting function may redistribute the entire probability mass of $\sum_i^n p(\omega_i) = 1$ across outcomes, the decision weights 
$$
\pi =
  \begin{cases}
     w(P(X \geq \omega_i)) - w(P(X > \omega_i)) \quad \text{for} \ 1, \\
     w(P(X \leq \omega_i)) - w(P(X < \omega_i)) \quad \text{else}, 
  \end{cases}
\tag{5}
$$
may 
people's deviations from EV maximization is often described in terms of the probability weighting pattern.

@@

```{r}
wf <- tibble(gamma = seq(.1, 1.9, .05)) %>%
  expand_grid(prob = seq(0, 1, .01)) %>%
  mutate(w = (prob^gamma)/((prob^gamma)+(1-prob)^gamma))

ggplot(wf) +
  geom_line(aes(prob, w, group = gamma, color = gamma)) +
  scale_color_scico(palette = "cork") + 
  scale_x_continuous(breaks = seq(0, 1, .1)) +
  scale_y_continuous(breaks = seq(0, 1, .1)) + 
  labs(x = expression(paste("p(", omega, ")")),
       y = expression(pi), 
       color = expression(gamma)) + 
  theme_apa() 
```

>>>>

Since there is an uncountable number of decisions people confront in the natural world, behavioral decision research routinely abstracts from decisions between particular choice alternatives, e.g., the choice between job offers, political parties to vote for, travel destinations, investment plans, and whatnot.
Rather, with the choice between at least two *prospects* (see below), it studies a case which is counterfactual in that it omits the many particularities of each decision, but retains in the form of prospects the fundamental *properties* that almost all choice alternatives are assumed to possess [@lopesThoughtsPsychologicalConcept1983].^[See @knightRiskUncertaintyProfit1921, Chapter 1, for a timeless discussion of the reasonableness of such abstractions.] 
These fundamental properties are considered to be the possible outcomes of a choice alternative---hereafter denoted by the Greek letter $\omega$---and the probabilities with which these outcomes occur following the choice of the alternative---hereafter $p(\omega)$.
That is, any choice alternative that can be fully described by its outcomes and their probabilities is a prospect. 
For illustration, take the choice alternatives in the following example with monetary outcomes:

@@

Now, let the exact axiomatic definition of a prospect be that of a *probability space* denoted by the tuple $(\Omega, \mathcal{F}, P)$. 
Following the @kolmogorovFoundationsTheoryProbability1950 axioms, $\Omega$ denotes the sample space, containing a finite set of possible outcomes:
$$
\omega_i \in \{\omega_1, ..., \omega_n\} = \Omega
\; .
\tag{1}
$$ 
$\mathcal{F}$ is the set of all possible subsets of $\Omega$, i.e., the power set $\mathcal{P}(\Omega)$ with each element $A \in P(\Omega)$ being a subset of $\Omega$.^[Note that this paper considers only the case of finite sample spaces. The exact axiomatic definition of $\mathcal{F}$, which can be found in Kolmogorov [-@kolmogorovFoundationsTheoryProbability1950, pp. 2-3 and 14-15], also covers infinite sample spaces and is that of a $\sigma$-algebra. For the finite case, however, $\mathcal{P}(\Omega)$ can be considered as such [@georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.1].] 
$P$ is a probability measure
$$
P: \mathcal{F} \mapsto [0,1]
\; ,
\tag{2}
$$
which assigns each subset $A \in \mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.^[The probabilities $p(\omega)$ are determined by the probability mass function $p : \Omega \mapsto [0,1]$, which is used to construct the probability measure $P(A) = \sum_{\omega \in A} p(\omega)$ for all $A \in \mathcal{F}$.]
Given a choice between 1, ..., k prospects, the inputs from the environment or memory should therefore provide information about the set
$$
I = \{(\Omega_1, \mathcal{F}_1, P_1), ... ,(\Omega_k, \mathcal{F}_k, P_k)\}
\; .
\tag{3}
$$

However, rarely in peoples' daily life, inputs take the form of explicit descriptions of all outcomes and their probabilities (as in Example 1).
In such cases, people would make *decisions from description* (DfD), since the information from which the mind learns about the choice alternatives---the probability spaces in (3)---are complete descriptions thereof.
Yet, rather often, people make *decisions from experience* [DFE, @hertwigDecisionsExperienceEffect2004], where the mind can learn about the choice alternatives only from a record of past experiences with them, which necessarily provides only incomplete and probable information.
That is, people may have experienced the possible outcomes of a choice alternative when they themselves or others have made the same decision repeatedly in the past, and the record of experienced outcomes for this particular choice alternative is denoted a *sample*. 
The mind may then infer from the relative frequency with which an outcome occurs in such a sample of past decisions its probability of occurrence for a future decision. 
Of course, all other things being equal, the accuracy of this inference depends on the size of the sample.
For illustration, take the sampled outcomes in the following example, where alternative A2 from Example 1 was chosen ten times without prior knowledge about the possible outcomes of this alternative and their actual probabilities:

@@

\noindent
One may now make an inference from the sample about the possible outcomes of choice alternative A2 and their probabilities. 
However, this conclusion is not certain, but only probable, since the relative frequencies of the results will almost surely change as the sample size increases.

@@

The choices people actually make between prospects---be that on the basis of ---are often described in terms of deviations from expected value (EV) maximization, where the EV is given by 
$$
EV = \sum_i^n \omega_i \times P(X = \omega_i)
\; .
\tag{6}
$$ 
That is, following the principle of EV maximization, one would first calculate the prospects' EV according to (6) and thereafter chose the prospect with the largest EV.
To *describe* how peoples' choices deviate from EV maximization [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations], the models of cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992; see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data with a combination of two parameterized functions that may replace the objective outcomes and probabilities in (6) by nonlinear transformations thereof.^[Note that CPT or other originally deterministic decision models may be amended by a stochastic choice function to account for the probabilistic nature of choices [see, e.g., @rieskampProbabilisticNaturePreferential2008].] 
That is, the models of CPT fit choice data by maximizing the value 
$$
V = \sum_i^n v(\omega_i) \times \pi_i
\; ,
\tag{7}
$$ 
with subjective values $v(\omega)$ derived from a *value function* that may transform the objective outcomes, and decision weights $\pi$ derived from the objective probabilities' cumulative distribution function that may be transformed by a *probability weighting function*. 
Now, since the probability weighting function may redistribute the entire probability mass of $P(\Omega) = 1$ across outcomes [@zilkerNonlinearProbabilityWeighting2021], the decision weights 
$$
\pi =
  \begin{cases}
     w(P(X \geq \omega_i)) - w(P(X > \omega_i)) \quad \text{for} \ \omega \geq 1, \\
     w(P(X \leq \omega_i)) - w(P(X < \omega_i)) \quad \text{else}, 
  \end{cases}
\tag{8}
$$
may be larger or smaller than the objective probabilities.

@@

```{r}
wf <- tibble(gamma = seq(.1, 2, .1)) %>%
  expand_grid(delta = c(.1, .5, 1, 2, 5, 10),
              prob = seq(0, 1, .01)) %>%
  mutate(w = (delta*prob^gamma)/((delta*prob^gamma)+(1-prob)^gamma))

ggplot(wf) +
  geom_line(aes(prob, w, group = gamma, color = gamma)) +
  geom_segment(x = 0, xend = 1, y =  0, yend = 1, linetype = "dashed") +
  scale_color_scico(palette = "cork") + 
  scale_x_continuous(breaks = seq(0, 1, .2)) +
  scale_y_continuous(breaks = seq(0, 1, .2)) + 
  labs(x = expression(paste("p(", omega, ")")),
       y = expression(pi), 
       color = expression(gamma)) + 
  facet_wrap(~delta, scales = "fixed") + 
  theme_apa() 
```
